---
title: "Final Project"
author: "Kyle Morris"
date: "May 29, 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(pastecs)
library(dplyr)

```
## Purpose

We are investigating here and running statistical analysis on soccer data. As a longtime sports fan, statistics are now the name of the game and I was interested in seeing how electronic players held up to their real world counterparts, and whether there is easily detected bias to the ratings systems which ultimately are objective.

We'll be using a number of packages in our analysis, including:

* ggplot2, a graphing package.
* pastecs, statistical analysis.
* dplyr, for data cleaning.

## Our Data

Our data comes to us from Kaggle. More specifically, we have three data sets we are looking at:

*	Kaggle – FIFA 19 Ultimate Team
	https://www.kaggle.com/stefanoleone992/fifa-19-fifa-ultimate-team
*	Kaggle – FIFA 19 Player Database
	https://www.kaggle.com/karangadiya/fifa19/
*	Kaggle – World Cup Players
	https://www.kaggle.com/djamshed/fifa-world-cup-2018-players

All of these are related to soccer (or Football, as those across the pond would insist on it being called.) The first two datasets are from FIFA 19 by Electronic Arts, the juggernaut that has seen over 20 million units sold to date alone. While it does face competition from Pro Evolution Soccer, FIFA is still the undisputed champion of the simulated soccer world.

Our last dataset concerns the 2018 World Cup. We were interested in how simulated soccer royalty compared to real world soccer royalty.

To begin with, let's import our data.

```{r}

fifaUltimate <- read.csv("fifa19ultimate.csv", header = TRUE)
fifaGame <- read.csv("fifagamedata.csv", header = TRUE)
worldCup <- read.csv("wc2018-players.csv", header = TRUE)
worldCup <- worldCup[-c(5)]
```

## Cleaning the data.

Some preliminary analysis of our data and the form it comes to us.

```{r}


fifaUltimateNA <- fifaUltimate[complete.cases(fifaUltimate), ]
fifaGameNA <- fifaGame[complete.cases(fifaGame), ]
worldCupNA <- worldCup[complete.cases(worldCup), ]

badFifaUltimate <- nrow(fifaUltimate) - nrow(fifaUltimateNA)
badFifaGame <- nrow(fifaGame) - nrow(fifaGameNA)
badWorldCup <- nrow(worldCup) - nrow(worldCupNA)

badFifaUltimate * 100 / nrow(fifaUltimate)
badFifaGame * 100 / nrow(fifaGame)
badWorldCup * 100 / nrow(worldCup)

```

For the ultimate data:

*	There is not a single complete entry in the entire 18831 rows of data. 
* That being said, there are 95 variables and the vast majority have almost every column. 
*	Missing fields are NA.

For the FIFA data:

*	Only 0.33% of the rows are missing any data. There are 18207 rows!
*	There are 89 variables tracked.
*	Missing results are marked NA. There are only 60 total incomplete rows.

For the World Cup data:

* Confirmed list of player names and birth date by country.
* Dataset has been curated based on FIFA records. This is the only complete data set we have – there is no missing data.
*	Data is from official rosters provided by FIFA.
*	No information on how it was originally collected. Small enough data set it could be done by hand.
*	There are 10 variables tracked and 736 observations.

## Preliminary Analysis

We will now take a look at scatterplots of our data.

```{r}

ggplot(subset(fifaUltimate, ps4_last > 0), aes(x = ps4_last, y = overall)) +
  geom_point(position = "jitter", alpha = 0.3) +
  geom_smooth(col = "red")
```

This graph charts the last sale price of the card on the PS4 versus the overall rating.

```{r}
ggplot(fifaUltimate, aes(x = age, y = overall)) +
  geom_point(position = "jitter", alpha = 0.3) +
  geom_smooth(col = "red")
```

This graph shows the Ultimate rating versus age. Some of the players are "legacy" in that they are famous players from the past and it shows the overall age.

```{r}


round(stat.desc(fifaUltimate$overall, basic = FALSE), digits = 3)

```

A look at our distribution of overall ratings.

```{r}

ggplot(fifaGame, aes(x = Age, y = Overall)) +
  geom_point(position = "jitter", alpha = 0.3) +
  geom_smooth(col = "red")
```

Same age versus overall plot, but this time for the FIFA stats.

```{r}
ggplot(fifaGame, aes(x = Age, y = Potential - Overall)) +
  geom_point(position = "jitter", alpha = 0.3) +
  geom_smooth(col = "red")
```

This graph is a measure of age versus room for growth i.e. $ potential - overall$.

```{r}
round(stat.desc(fifaGame$Overall, basic = FALSE), digits = 3)

bw <- 2 * IQR(worldCup$Age) / length(worldCup$Age)^(1/3)
ggplot(worldCup, aes(x = Age)) +
  geom_histogram(binwidth = bw)
```

A histogram of the ages of players in the most recent World Cup.

```{r}
round(stat.desc(worldCup$Age, basic = FALSE, norm = TRUE), digits = 3)

cleanFifaGame <- subset(fifaGame, select = c("Name", "Age", "Overall", "Potential", "Club","Position", "Height", "Weight", "Crossing", "Finishing", "HeadingAccuracy" , "ShortPassing" , "Volleys" , "Dribbling" , "Curve", "FKAccuracy", "LongPassing" , "BallControl" , "Acceleration" , "SprintSpeed" , "Agility" , "Reactions", "Balance" , "ShotPower" , "Jumping" , "Stamina" , "Strength" , "LongShots", "Aggression" , "Interceptions" , "Positioning", "Vision" , "Penalties" , "Composure", "Marking" , "StandingTackle" , "SlidingTackle" , "GKDiving" , "GKHandling", "GKKicking", "GKPositioning", "GKReflexes"))
cleanFifaGameNA <- cleanFifaGame[complete.cases(cleanFifaGame), ]

head(cleanFifaGameNA)

```

We've now cleaned up the FIFA stats as those will be the most useful. We kept just the most useful fields for our analysis. Now, we can begin!

## What are some of the questions we might be looking to answer?

Electronic Arts has done their best to score over 19000 football players around the world. On their platform, all skills should in theory be ranked accordingly. The Overall score is an easy catch-all explanation of player skill, but how is that calculated? Is it the average of all scores? Is there a league that has, in general, a stronger overall player base than any other? If I was utilizing their Career mode, I would want to know the beginning set up that gave my future superstar the highest overall potentional!

In order to address that,, we should look to address the following: which league has the highest overall skill rating, how exactly does FIFA weight the various skill measurements to determine overall rating, and how we can maximize the potential of our new recruits.

## Our Model

Let's create our model. I'm going to split the data at this point as well, because I suspect that the model for goalkeepers is going to look significantly different than the model for all players, or just other position players.

```{r, warning = FALSE}
cleanGK <- subset(cleanFifaGameNA, cleanFifaGameNA$Position == "GK")
cleanOthers <- anti_join(cleanFifaGameNA, cleanGK)

gkTest <- cleanGK %>% sample_frac(.2)
gkTrain <- anti_join(cleanGK, gkTest)

offenseTest <- cleanOthers %>% sample_frac(.2)
offenseTrain <- anti_join(cleanOthers, offenseTest)

overallModel <- glm(Overall ~ Age + Crossing + Finishing + HeadingAccuracy + ShortPassing + Volleys + Dribbling + Curve + FKAccuracy + LongPassing + BallControl + Acceleration + SprintSpeed + Agility + Reactions + Balance + ShotPower + Jumping + Stamina + Strength + LongShots + Aggression + Interceptions + Positioning + Vision + Penalties + Composure + Marking + StandingTackle + SlidingTackle + GKDiving + GKHandling + GKKicking + GKPositioning + GKReflexes, data = cleanFifaGameNA)

gkModel <-glm(Overall ~ Age + Crossing + Finishing + HeadingAccuracy + ShortPassing + Volleys + Dribbling + Curve + FKAccuracy + LongPassing + BallControl + Acceleration + SprintSpeed + Agility + Reactions + Balance + ShotPower + Jumping + Stamina + Strength + LongShots + Aggression + Interceptions + Positioning + Vision + Penalties + Composure + Marking + StandingTackle + SlidingTackle + GKDiving + GKHandling + GKKicking + GKPositioning + GKReflexes, data = gkTrain)

offenseModel <- glm(Overall ~ Age + Crossing + Finishing + HeadingAccuracy + ShortPassing + Volleys + Dribbling + Curve + FKAccuracy + LongPassing + BallControl + Acceleration + SprintSpeed + Agility + Reactions + Balance + ShotPower + Jumping + Stamina + Strength + LongShots + Aggression + Interceptions + Positioning + Vision + Penalties + Composure + Marking + StandingTackle + SlidingTackle + GKDiving + GKHandling + GKKicking + GKPositioning + GKReflexes, data = offenseTrain)

gkTest$model_pred <- round(predict(gkModel, gkTest, type = "response"))

offenseTest$model_pred <- round(predict(offenseModel, offenseTest, type = "response"))

ggplot(gkTest, aes(x = Overall, y = model_pred)) +
  geom_jitter() +
  geom_smooth(aes(color = "red"))
```

Wow, our model for goalkeepers looks really, really nice. There are a few outliers but for the most part the model looks very good at predicting the overall grade. It looks like there might not be as much nudge factor as expected. 

Here's the same graph but for the test set of regular players:

```{r, warning = FALSE}

ggplot(offenseTest, aes(x = Overall, y = model_pred)) +
  geom_jitter() +
  geom_smooth(aes(color = "red"))+
  ylab("Predicted Overall")

```

Overall a pleasing graph. I added some jitter due to the number of datapoints but generally there is an overall trend.

What variables are most significant? We have all of our data on the same scale so this analysis is slightly easier.

```{r}
summary(gkModel)

summary(offenseModel)

```
For goalkeepers, there are 5 variables that have quite an effect on the overall score:

1. GKHandling
2. GKDiving
3. GKReflexies
4. GKPosition
5. Reactions

For regular players, it is slightly different:

1. Reactions
2. Ball Control
3. Composure
4. Short Passing
5. Heading Accuracy

Reactions is important for both sets of players, but that is it. 
```{r}
gkTest <- mutate(gkTest, differential = (model_pred - Overall))
offenseTest <- mutate(offenseTest, differential = (model_pred - Overall))

overachievingOffense <- subset(offenseTest, differential < 0)
overachievingGK <- subset(gkTest, differential < 0)

underachievingOffense <- subset(offenseTest, differential > 2)
underachievingGK <- subset(gkTest, differential > 2)


```

There are `r nrow(overachievingGK)` Goalkeepers who had ranked slightly higher than the model predicts and `r nrow(overachievingOffense)` regular players. this corresponds to `r round(((nrow(overachievingGK) / nrow(gkTest)) * 100), digits = 3)`% of the goalkeepers and `r round(((nrow(overachievingOffense) / nrow(offenseTest)) * 100), digits = 3)`% of the regular players. Yikes! However, if we look at it a different way:

```{r}
overachievingOffense.2 <- subset(overachievingOffense, abs(differential) >= 2)
overachievingGK.2 <- subset(overachievingGK, abs(differential) >= 2)

overachievingOffense.5 <- subset(overachievingOffense, abs(differential) >= 5)

```

There are now `r nrow(overachievingGK.2)` goalkeepers that are 2 or more points away from their expected, and `r nrow(overachievingOffense.2)` regular players. So it seems goalkeepers are spot on but players are a little trickier! 

However, `r nrow(overachievingOffense.5)` players are 5 or more points higher than the model predicts.

So who are the players we would find most desirable? Those would be the players that have an overall rating higher than our model predicts. They are more than the sum of their parts, basically.

We'll now look at the top 6 players per position, based on them being better than the model predicts.

```{r}


overachievingOffense <- overachievingOffense[order(overachievingOffense$differential),]

overachievingGK <- overachievingGK[order(overachievingGK$differential),]

print("Central Attacking Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "CAM"), Name, Overall, differential))
print("Center Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "CB"), Name, Overall, differential))
print("Central Defensive Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "CDM"), Name, Overall, differential))
print("Center Forward:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "CF"), Name, Overall, differential))
print("Center Midfield:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "CM"), Name, Overall, differential))
print("Left Attacking Midfield:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LAM"), Name, Overall, differential))
print("Left Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LB"), Name, Overall, differential))
print("Left Center Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LCB"), Name, Overall, differential))
print("Left Center Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LCM"), Name, Overall, differential))
print("Left Forward:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LF"), Name, Overall, differential))
print("Left Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LM"), Name, Overall, differential))
print("Left Striker:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LS"), Name, Overall, differential))
print("Left Wing:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LW"), Name, Overall, differential))
print("Left Wing Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "LWB"), Name, Overall, differential))
print("Right Attacking Midfield:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RAM"), Name, Overall, differential))
print("Right Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RB"), Name, Overall, differential))
print("Right Center Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RCB"), Name, Overall, differential))
print("Right Center Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RCM"), Name, Overall, differential))
print("Right Forward:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RF"), Name, Overall, differential))
print("Right Midfielder:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RM"), Name, Overall, differential))
print("Right Striker:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RS"), Name, Overall, differential))
print("Right Wing:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RW"), Name, Overall, differential))
print("Right Wing Back:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "RWB"), Name, Overall, differential))
print("Striker:")
head(select(subset(overachievingOffense,overachievingOffense$Position == "ST"), Name, Overall, differential))
head(select(overachievingGK, Name, Overall, differential))

```

## Analysis

Overall, how accurate was our model? Here we will define inaccuracy as +- 2 points from predicted.

For regular players:

```{r}
accuracyOffense <- (1 - ((nrow(underachievingOffense) + nrow(overachievingOffense.2)) / nrow(offenseTest))) * 100
```

For goalkeepers:

```{r}
accuracyGK <- (1 - ((nrow(underachievingGK) + nrow(overachievingGK)) / nrow(gkTest))) * 100
```

The model is accurate for 96.3% of Goalkeepers in our test set, and 58.6% of players in our regular set. Still, when one considers the implication none of our players are more than 9 points off from prediction so it is actually a pretty decent model despite being off by 2 points for 40% of players.

## Conclusions

Based on our analysis, the overall skill is a fairly linear relationship between the different skills measured in our players. While there are some outliers, given a set of skills we can reasonably predict within a few points where that player will lie. While the initial analysis of players may have its own set of biases, once a player is graded their skills are weighted equally across the board. Goalkeepers in particular adhere fairly strictly to this. 

We have now identified the top 5 skills for both sets of players in order to maximize your overall gains.These are the skills you should focus on to have the greatest overall effect.

This analysis is solely limited to FIFA 19. We may be able to improve the accuracy of the model by controlling for player position, something that we did not do other than by splitting out the goalkeepers.

Overall, I was impressed with how much of a relationship existed that was quantifiable. I was worried initially that there would be some liberties taken with scoring. I was pleasantly surprised to find that was not the case.

